---
title: "APackOfTheClones Essentials"
description: >
  A brief walkthrough of the clonal expansion visualization workflow.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APackOfTheClones Essentials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
data: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_chunk$set(echo = FALSE)
options(repos = c(CRAN = "http://cran.rstudio.com"))

# utility functions
head <- function(df) {
  knitr::kable(utils::head(df))
}

quiet_load_all_CRAN <- function(...) {
  for (pkg in list(...)) {
    if (require(pkg, quietly = TRUE, character.only = TRUE)) next
    invisible(install.packages(
      pkg, quiet = TRUE, verbose = FALSE, character.only = TRUE
    ))
    suppressPackageStartupMessages(invisible(
      require(pkg, quietly = TRUE, character.only = TRUE)
    ))
  }
}

# load packages
quiet_load_all_CRAN("ggplot2", "cowplot", "Seurat")

# TODO improve entire vignette
```

```{r setup}
suppressPackageStartupMessages(library(APackOfTheClones))
```

## Introduction

Single-cell RNA sequencing (scRNA-seq) and T/B cell receptor (TCR) sequencing are popular techniques for studying immune cell function and disease. The combined use of such data can provide valuable insights into the immune system, including clonal expansion. `APackOfTheClones` provides a simple, easily customizable, and publication-ready method to intuitively visualize clonal expansion between different cell clusters with `ggplot2`, and can be easily slotted into any analysis pipeline.

The method counts clonotypes, and using a dimensional reduction (e.g. UMAP) of all cells in a single cell immune profiling experiment as a base, it circle-packs all clone sizes directly as circles within circular clusters according to its seurat cluster. The advantage of this method compared to other (very limited) visualizations of clonal expansion out there is that its arguably much more intuitive. To see it in action, see [this paper](https://doi.org/10.1126/sciimmunol.abg6356) and [this paper](https://doi.org/10.1111/all.15399).

Basic familiarity with the `R` language, the `Seurat` package, and the `scRepertoire` package *VERSION TWO* is assumed. As version two of `scRepertoire` is relatively new and introduces some small breaking changes, it is *essential* that its [new vignettes](https://www.borch.dev/uploads/screpertoire/) are read.

In this vignette, the most essential functionalities of the package are covered, from preparing the data with scRepertoire, to producing and fine-tuning a barebones visualization of clonal expansion.

### Setting up the Seurat Object and Receptor Library with scRepertoire

The premise of the package is that it provides an additional analysis tool on top of `scRepertoire`'s many functions, which are all ran after combining all TCR/BCR contigs into clones, and integrating the clonal information into a seurat object with a dimensional reduction(s). Here are the corresponding vignettes to read in order for combining contigs:

1. [Loading Data](https://www.borch.dev/uploads/screpertoire/articles/loading)
2. [Combining Contigs into Clones](https://www.borch.dev/uploads/screpertoire/articles/combining_contigs)
3. [Additional Processing Steps (optional)](https://www.borch.dev/uploads/screpertoire/articles/processing)

And here is a practical example on how to generate a combined seurat + VDJ object to be used in the vignette named `pbmc`, which is also identical to the built-in example dataset `combined_pbmc` which can be loaded with `data("combined_pbmc", package = "APackOfTheClones")`:

```{r, combineTCR, eval = FALSE}
library(scRepertoire)

# load in the corresponding 6-sample TCR contigs from scRepertoire
contig_list <- get(data("contig_list", package = "scRepertoire"))

# combine the TCR contigs into clones with custom samples
combined_contig_list <- scRepertoire::combineTCR(
  contig_list,
  samples = c("P17B", "P17L", "P18B", "P18L", "P19B", "P19L", "P20B", "P20L"),
  removeNA = FALSE, 
  removeMulti = FALSE, 
  filterMulti = FALSE
)
```

For integrating the clonal information, ```scRepertoire::combineExpression``` is the function used [to do so](https://www.borch.dev/uploads/screpertoire/articles/attaching_sc), but for `APackOfTheClones` the scRNA-seq object has to be a seurat object:

```{r, combining, eval = FALSE}
# a seurat object corresponding to combined_contig_list named `pbmc` is loaded
pbmc <- scRepertoire::combineExpression(
  combined_contig_list,
  pbmc,
  cloneCall = "gene",
  proportion = TRUE
)

print(pbmc)
```

```{r, actual_print_pbmc, eval = TRUE, echo = TRUE, include = FALSE}
# TODO use a nicer looking dataset
pbmc <- get(data("combined_pbmc"))
print(pbmc)
```

And as a reference, here is the corresponding UMAP plot of the seurat object

```{r, umap, echo = TRUE}
library(Seurat)
library(ggplot2)

pbmc_umap_plot <- UMAPPlot(pbmc)
pbmc_umap_plot + coord_fixed()
```

## Simple ball packing visualization

```vizAPOTC``` (short for "visualize APackOfTheClones") is the main convenience function of the package to directly produce the ball packing clonal expansion plot. It takes in a main argument of a combined seurat object (with a long list of optional arguments which will be covered in later sections), and outputs a ggplot object:

```{r, initial_vizapotc, echo = TRUE}
default_apotc_plot <- vizAPOTC(pbmc, verbose = FALSE)
default_apotc_plot
```

### Key characteristics of the plot

* One should notice the immediate correspondence of circular clusters to umap clusters, and be able to derive some immediate insights about which clusters have expanded.
* The clonotype counts for each seurat cluster corresponds to their position and color in the original UMAP centroids
* The most expanded clonotypes are in the center of each circle cluster with larger sizes, symbolizing increased expansion
* On the plot, there is a somewhat imperfect visual legend of the relative clone sizes
* Some clusters have considerable visual overlap
* The returned plot is a fully customizable `ggplot` object

The resulting clonal expansion plot may not be visually satisfactory on the first run without customizations from optional arguments. These arguments and other `ggplot` tricks will be covered to fine-tune the visualization until publication-ready.

## Working with different reduction bases and clonal representations

The arguments to do so are:

```{r, echo = TRUE, eval = FALSE}
reduction_base = "umap",
clonecall = "strict",
```

`reduction_base` indicates what each seurat cluster's centroid locations should be based upon, so UMAP, T-SNE, or PCA (provided the reduction has been ran on the object). `clonecall` corresponds to which column in the seurat metadata should be used to conduct the clonotype counting. These are usually columns generated by combineTCR/BCR but can be custom columns too. Note that ```scRepertoire``` should generate the following columns by default, and the brackets indicate equivalent names one can pass into the clonecall parameter:

- CTgene (gene, genes)
- CTnt (nt, nucleotide, nucleotides)
- CTaa (aa, amino)
- CTstrict (strict, gene+nt)

Switching this parameter could be useful if the user had generated a custom definition of clones and added it to the meta data, or if BCR clonal data is present, since [the definition of a clone for B-cells isn't as clear as it is for TCRs](https://doi.org/10.3389/fimmu.2023.1123968).

## Working With a Data Subset

A novel feature of version 1 is the ease of running APackOfTheClones for subsets of the full dataset, which may be useful in cases like the need for clonal expansion plotting for only certain samples or conditions. There are two arguments to do so:

```{r, echo = TRUE, eval = FALSE}
...,
extra_filter = NULL,
```

`...` represents an arbitrary number of additional keyword arguments indicating the rows corresponding to elements in the seurat object metadata that should be filtered by. For example, seurat_clusters = c(1, 9, 10) will filter the cells to only those in seurat clusters 1, 9, and 10. `extra_filter` is another additional way to subset the data and should be formatted *exactly* like a statement one would pass into `dplyr::filter` that does *additional* filtering to cells in the seurat object.

Here is an example of the data subsetting, where only samples in the example seurat object that correspond to sample 17, 18, and 19 are plotted, alongside the original plot to see the difference. Optionally, it may sometimes be helpful to set one of the visual parameters `retain_axis_scales` to `TRUE`, extending the output plot's dimensions if needed to approximately match the corresponding reduction plot's dimensions:

```{r, subsetting, echo = TRUE, fig.dim = c(6, 6)}
# `orig.ident` is a custom column in the example data with levels corresponding to sample ids:
# ("P17B" "P17L" "P18B" "P18L" "P19B" "P19L" "P20B" "P20L"). Here, it is subsetted
# by the keyword argument approach
subset_sample_17_plot <- vizAPOTC(
  pbmc,
  orig.ident = c("P17B", "P17L"),
  retain_axis_scales = TRUE,
  add_size_legend = FALSE,
  verbose = FALSE
)

# here, it is subsetted with `extra_filter` for sample 18 with dplyr syntax:
subset_sample_18_plot <- vizAPOTC(
  pbmc,
  extra_filter = "substr(orig.ident, 1, 3) == 'P18'",
  retain_axis_scales = TRUE,
  add_size_legend = FALSE,
  verbose = FALSE
)

# here, sample 19 is subsetted with both arguments to show that they work in conjunction
subset_sample_19_plot <- vizAPOTC(
  pbmc,
  orig.ident = "P19B",
  extra_filter = "orig.ident == 'P19L' | orig.ident == 'P19B'",
  retain_axis_scales = TRUE,
  add_size_legend = FALSE,
  verbose = FALSE
)

cowplot::plot_grid(
  vizAPOTC(combined_pbmc, add_size_legend = FALSE, verbose = FALSE),
  subset_sample_17_plot,
  subset_sample_18_plot,
  subset_sample_19_plot,
  labels = c("all", "17", "18", "19")
)
```

### A Note on Subsetted Data

The parameters are barely more than syntactic sugar to subset the seurat metadata and dimensional reduction coordinates, and serves as a convenience to the user if they want to quickly generate insights. Generally, if the analysis of data subsets (such as certain samples, or condition, etc.) is crucial to a publication, one would likely already be working with fully subsetted seurat object, and APackOfTheClones functions directly on those. See the [corresponding `Seurat` `vignette("essential_commands")`](https://satijalab.org/seurat/articles/essential_commands.html#subsetting-and-merging) for details.

## Customization of Visual Parameters

### Visually Scaling Circle Sizes

For each seurat cluster at index $i$, the final radii $r_{ij}$ of each physical circle at index $j$ representing clone size $s_{ij}$ is calculated with a clone size scaling factor $C \in [0, 1)$ and a radius scaling factor $R \in [0, 1)$ with the following formula:

$$r_{ij} = C\cdot(\sqrt{s_{ij}} - (1 - R))$$

Intuitively, $C$ represents how much to enlarge/shrink each radius geometrically, whereas $R$ represents the scaled size of the smallest clone's radius, and all circles will have their radii decreased by that amount. $R$ defaults to 0.95, whereas $C$ defaults to an approximated factor based on the number of clones that the package computes. The corresponding arguments are `clone_scale_factor` and `rad_scale_factor` in `vizAPOTC`.

### Cluster Repulsion

Considerable visual overlap between clusters (due to the algorithm's attempt to fit clusters to the original UMAP coordinates) may occur and obstruct eachother excessively, even though the method by default tries to alleviate this with repulsion of clusters. There are four optional arguments:

```{r, echo = TRUE, eval = FALSE}
repulse = TRUE,
repulsion_threshold = 1,
repulsion_strength = 1,
max_repulsion_iter = 10
```

* `repulsion_threshold` indicates the amount of `ggplot2` units of overlap between clusters that are acceptable. It defaults to `1`, meaning that two clusters that overlap by about 1 unit are considered by the repulsion algorithm to not be overlapping. Increasing this number will increase the amount of overlap between clusters, and decreasing this number will do the opposite, while decreasing it to negative values will incur additional spacing between clusters. However, using negative spacing may cause the plot to look very spaced out. Keep reading to see alternatives to doing so.
* `repulsion_strength` relates to how much the clusters should repel each other. The repulsion algorithm works in iterations, where for each iteration, each cluster "pushes" each other away from eachother by some amount. Increasing this value will cause extra "pushing" during each iteration. However, increasing this factor too much may result once again in a very visually unpleasant plot.
* `max_repulsion_iter` indicates the number of iterations where clusters should repel eachother. Increasing this number would ensure that clusters will (almost always) for sure not be overlapping. A trick with this parameter to make more pleasant plots is to decrease `repulsion_strength` and increase `max_repulsion_iter` to possibly make a more pleasant arrangement of clusters.

For more details on them, read the "Arguments" section in the function documentation. Briefly, first, to make the circle clusters move away from eachother, `repulse` should be set to `TRUE`, and the function should be ran AGAIN. (If you feel this excessive re-running takes too long or is inefficient for your workflow in its current form due to a constant need to fine-tune, see the vignette `APackOfTheClones-runs`.)

### Managing the Clone Size Legend

There are the following six parameters to adjust the legend:

```{r, legend_params, echo = TRUE, eval = FALSE}
add_size_legend = TRUE,
legend_sizes = "auto",
legend_position = "auto",
legend_buffer = 0.2,
legend_color = "#808080",
legend_spacing = "auto",
legend_label = "Clone sizes",
legend_text_size = 5,
add_legend_background = TRUE,
```

`legend_sizes` are autogenerated, with at least sizes for clone size 1 and max(clone sizes). The user can also input whichever sizes they wish as a numeric vector. The `legend_position` on the plot defaults to one of the four corners, whichever will result in the least overlap of the legend with circles present. Otherwise, the user can input manually `"top left"`, `"top right"`, `"bottom left"` `"bottom right`, or just a numeric indicating the x and y coordinates of the top leftmost corner of the legend. `add_legend_background` will display a rectangular background to the legend. The other parameters are less relevant and more details about these arguments can be read in the function documentation.

### Other Modifications

A collection of other utility parameters are:

```{r, other_params, echo = TRUE, eval = FALSE}
order_clones = TRUE,
try_place = FALSE,
res = 360L,
linetype = "blank",
use_default_theme = TRUE,
retain_axis_scales = FALSE,
show_labels = FALSE,
label_size = 5,
```

The most important of which are 1. `show_labels` which overlays the seurat cluster on the plot in the format `"Cx"` in the center of each relevant cluster. (see `vignette("APackOfTheClones-runs")` on how one can modify them) 2. `use_default_theme` which will produce a plot that's thematically very similar to the dimensional reduction plot, with axis labels and numeric ticks on each axis. This may be helpful for visually indicating certain plot parameters. However, for publications [such as this one](https://doi.org/10.1126/sciimmunol.abg6356), as the numeric ticks in APackOfTheClones have no actual biologically relevant meaning aside from approximately matching clusters, it may be beneficial to set this argument to `FALSE` which will produce a plot with just the circles:

```{r, void_labelled_plot, echo = TRUE}
vizAPOTC(
  pbmc,
  show_labels = TRUE,
  label_size = 3,
  legend_text_size = 3.5,
  add_legend_background = FALSE,
  use_default_theme = FALSE,
  verbose = FALSE
)
```

## Conclusion

That's about it for the most basic functionalities of the clonal expansion visualization function. It's ***strongly recommended*** to save the plot first as an `.svg` file with ```ggplot2::ggsave``` for maximal resolution (especially of the circles) for publication.

For users that need to fine-tune plot parameters and/or save the data within the seurat object for readjustment/replotting, please read the `vignette("APackOfTheClones-runs")`.

For inspiration of how it could practically look like in a real paper context, see the following papers where the original julia implementation of APackOfTheClones was successfully used:

* [Single-cell analysis pinpoints distinct populations of cytotoxic CD4+ T cells and an IL-10+CD109+ $T_{H}2$ cell population in nasal polyps](https://doi.org/10.1126/sciimmunol.abg6356)
* [Recombinant multimeric dog allergen prevents airway hyperresponsiveness in a model of asthma marked by vigorous $T_{H}2$ and $T_{H}17$ cell responses](https://doi.org/10.1111/all.15399)
