---
title: "Working With Clonotypes Shared Across Clusters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working With Clonotypes Shared Across Clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
data: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_chunk$set(echo = FALSE)
options(repos = c(CRAN = "http://cran.rstudio.com"))

quiet_load_all_CRAN <- function(...) {
  for (pkg in list(...)) {
    if (require(pkg, quietly = TRUE, character.only = TRUE)) next
    invisible(install.packages(
      pkg, quiet = TRUE, verbose = FALSE, character.only = TRUE
    ))
    suppressPackageStartupMessages(invisible(
      require(pkg, quietly = TRUE, character.only = TRUE)
    ))
  }
}

# load packages
quiet_load_all_CRAN("ggplot2", "Seurat", "magrittr", "viridis", "APackOfTheClones")

# load data
pbmc <- get(data("combined_pbmc"))
```

## Introduction

A novel experimental feature in this package is the ability to get, filter, and visualize clonotypes that are common across seurat clusters. Potential applications include the ability to quickly and intuitively gauge the rough clonal heterogeneity of certain clusters. Note that in scRepertoire, the `scRepertoire::clonalNetwork` function does something similar for the original dimensional reductions using arrows, though those display the aggregated degree of commonality between each cluster.

Please read `vignette("APackOfTheClones")` and ideally also `vignette("APackOfTheClones-runs")` before reading this vignette.

## Getting Clonotypes Common Across Seurat Clusters

`getSharedClones()` is a convenience function does so while allowing the same seurat object subsetting arguments as seen in `vignette("APackOfTheClones")`, and returns the clonotypes as a named list where the *names* are the clonotypes and the *elements* are numeric vectors where each value is one of the clusters that the clonotype name at its index correspond to.

```R
getSharedClones(
    seurat_obj,
    reduction_base = "umap",
    clonecall = "strict",
    ...,
    extra_filter = NULL,
    run_id = NULL,
    top = NULL,
    top_per_cl = NULL
)
```

The last two arguments help allow additional filtering of the clonotypes: `top` and `top_per_cl`. The former will filter the clonotypes to only those ranked in the `top` frequencies, either as a proportion or number, while the latter does so for each individual cluster. These arguments are especially useful when an experiment has a really high number of single-cells, and the only clonotypes of interest are the most expanded ones.

Here is an example of the function in action, where a combined seurat object is loaded with the variable name `pbmc`:

```{r, shared_clones, echo = TRUE}
head(getSharedClones(pbmc, clonecall = "aa"))
```

## Visualizing Shared Clonotypes on an APackOfTheClones Plot

Using the output from `getSharedClones()`, `vizAPOTC()` and `APOTCPlot()` both currently have the exact same arguments to take the information and visually overlay line links between each shared clone. Note that many more changes are to come to this functionality. The arguments are as follows:

```R
show_shared = NULL,
only_link = NULL,
clone_link_width = "auto",
clone_link_color = "black",
clone_link_alpha = 0.5
```

Where if the output of `getSharedClones` is inputted into `show_shared`, the links will be overlaid. The `only_link` argument will further filter the links so that all links originate from the cluster `only_link`. The other are aesthetic arguments for the lines and are self explanatory. More details in the function level docs.

Here is an example, using `pbmc`:

```{r, shared_clones_links, echo = TRUE}
# get shared amino acid shared clones:
shared_clones_aa <- getSharedClones(pbmc, clonecall = "aa")

# generate the plot
vizAPOTC(
    pbmc,
    clonecall = "aa",
    show_shared = shared_clones_aa,
    verbose = FALSE
)
```

A more useful visualization may be to overlay the links from one cluster, and setting `clone_link_color` to `"blend"` which will color each clone link according to the average color of two clusters, for even more intuitiveness:

```{r, shared_clones_links_from_1_blend, echo = TRUE}
vizAPOTC(
    pbmc,
    clonecall = "aa",
    show_shared = shared_clones_aa,
    only_link = 4, # only link clonotypes from cluster 4
    clone_link_color = "blend",
    clone_link_alpha = 0.8,
    verbose = FALSE
)
```

### Synergy with `showCloneHighlight()`

In the interest of inspecting the topmost expanded clonotypes, the topmost clones can be obtained, and the plot of clonal links can have its shared clones highlighted with a different palette than the default ggplot hues, and the non-highlighted sequences can even be slightly dimmed to emphasize the difference. The following is an example of highlighting and linking the top 4 most expanded clonotypes in `pbmc`:

```{r, shared_clones_links_highlight, echo = TRUE}
# get shared amino acid shared clones:
shared_clones_aa_top4 <- getSharedClones(pbmc, clonecall = "aa", top = 4)

# generate the plot
linked_apotc_plot <- vizAPOTC(
    pbmc, clonecall = "aa",
    show_shared = shared_clones_aa_top4
)

# highlight the top 4 clones with the viridis palette
library(viridis)
showCloneHighlight(
    linked_apotc_plot,
    sequence = names(shared_clones_aa_top4),
    color_each = viridis(4),
    default_color = NULL,
    scale_bg = 0.75
)
```

Note that to actually get the frequencies of those top clones, one can use the `countCloneSizes` function.
